\documentclass{amsart}

\input{decl-db}

\begin{document}

\title{Categorical foundations for Ontologies and databases}

\author{David I. Spivak}

\maketitle

\tableofcontents

\section{Introduction}

Converting between databases and ontologies.  First: what are they??  Kinds of information, links between them, answers to queries.  Namely a functor from a category to sets.  What kind of category?

Category theory as a way to compare various models.  Also beneficial to force a coherent viewpoint.

The model we present.  Its flexibility w.r.t. inputs, instances as objects.  Colimits and limits, maps from "child" to "person" give the children of a person.  ``Dimensions" of an object $F$ are the ways to measure it, i.e. the set of objects of $\mcC$ which it maps to.

Rigorous definitions for: ontology, field, domain, recortd, data.

Transitivity enforced.

Flexible: High-end mathematics.

Advantages: Don't need disjoint data to make sense of this.  Like doors should be considered parts of two rooms: overlap.

\section{Ontologies}

The picture we draw and their defects.

Andrea's problem.

\section{Necessary category theory}

Basics.

The ``map of categories" viewpoint: i.e. taking a functor $F\taking\mcC\to\Sets$ and replacing it by a map $\mcD_F\to\mcC$ with sets as fibers.  We call this the fiber-view.

Adding a functor $\mcC\to\Sets$ to $\mcC$ (by taking the full subcategory of $\Fun(\mcC,\Sets)\op$ spanned by $\Ob(\mcC)\amalg \{F\}$. 

Pro-objects.

\section{Ontologies as Grothendieck sites}

Why it's appropriate.  

\section{Advantages and Applications}

Databases inside of databases.  (Math departments have people, and people are their own databases).

Category theoretic approach forces data consistency.

Physics: Schrodinger's cat -- an example where we can't decide what element of the cover we are in.

Adding more covering sieves (this product sold `well'.  His temperature was ``high.')

Puts data types, fields, values on the data, and entities on the same playing field (though entities can be made more ethereal by considering them as functors).

The conversion problem.

Data level manipulations: using the fiber-view of functors.  Drawing pictures.

Andreas problem.

Mathematical implications.  Describing mathematics.  Topological spaces as databases.

Linguistics? 

\end{document}